Scientific Reasoning: \\ Assessment of Multimodal Generative LLMs

### 2024-12-28 docs/paper/main.tex

\section{Metrics}
We evaluate model's performance with question answering accuracy domain-wise in order to have a fair comparison with leaderboards. The reasoning steps can be evaluated with semantic similarity metrics (e.g. adopted from machine translation) such as BLEU, METEOR, ROUGE, and cosine similarity and more.
\subsection{Multiple choice Evaluation}
Owing to the simplicity of the test format, only accuracy score is computed, following original evaluation strategy by \cite{lu2022learn}. It was determined that the computation of the overall score should assign the highest weight (\textbf{50\%}) to the ability to provide correct answers (accuracy score), as this represents the most critical aspect of performance evaluation.
\subsection{Answer Reasoning Evaluation}
Due to the pecularity of scientific texts and approaches to the evaluation of automatically generated texts, the following evaluation approaches were chosen. Our validation process is consisted of the metrics introduced in the original paper \cite{lu2022learn}  with a few additional metrics and techniques that we consider of high importance in regards to the scientific textual data.

@@

%\begin{example}
\label{example:1}
\begin{itemize}
    \item \textbf{Reference:} Plants absorb sunlight to produce energy.
    \item \textbf{Candidate:} Sunlight is absorbed by plants to create energy.
\end{itemize}
%\end{example}

\paragraph{Cosine similarity with Sentence Transformers}
Cosine similarity using SentenceTransformer embeddings provides a robust method for evaluating model outputs in the ScienceQA dataset, focusing on the semantic alignment of generated explanations with references. It captures nuanced relationships and conceptual understanding, which gives more understanding of the generated outputs in comparison to ROUGE and BLEU scores and that is why it has more importance in final score computation. However, due to domain specificity it can miss some intricate cases like still considering high similarity of two sentences, even though it the reference can contain crucial information about carbon dioxide, while in the candidate sentence the answer was given in relation to oxygen reproduction.
\paragraph{METEOR}
METEOR (\cite{banerjee2005meteor}) evaluates text similarity based on precision, recall, and alignment of semantically meaningful components like synonyms and stems, making it well-suited for capturing the nuanced language used in scientific reasoning.
By incorporating a weighted F-score and penalties for excessive mismatches, METEOR provides a balanced evaluation of fluency and accuracy, crucial for assessing LLM-generated answers on the Science QA Dataset. Due to these balanced text evaluation, this metric has \textbf{15\%} impact on the overall score.
% \subsubsection{Cosine similarity with SciBERT}
% We extended the evaluation of cosine similarity using SciBERT (\cite{beltagy2019scibert}) embeddings. These vectors are more consistent with the domain-specific terminology and reasoning patterns present in the Science QA Dataset because it has been properly pre-trained on scientific literature.
% SciBERT generates more accurate similarity embeddings because it better reflects the complex relationships between scientific concepts than general-purpose models like those in Sentence Transformers.
% When assessing the logical and factual coherence of scientific reasoning, the structural and contextual semantics specific to scientific texts are reflected in the embeddings produced by SciBERT.
% Assessments for jobs requiring sophisticated scientific understanding are more reliable when SciBERT is used for cosine similarity since it guarantees that the evaluation metric is domain-specific.

@@

We evaluate: 
\begin{enumerate}
	\item OpenAI models: \texttt{GPT-4}, \texttt{GPT-4o-mini}
	\item MistralAI model: \texttt{Pixtral-12b-2409}
    \item Gemini models: \texttt{Gemini-1.5-flash}, \texttt{Gemini-1.5-flash-8B}
	\item Llava model:  \texttt{llava-hf/llava-1.5-7b-hf}
\end{enumerate}

\subsection{Metrics}
% description of metrics, why they were used + analytics of results
\paragraph{Accuracy}
The highest performance was observed in settings that included solutions, indicating that the models were generally capable of extracting relevant information effectively. Among these, settings combining both lecture and solution information typically yielded slightly better results than those with solutions alone. \texttt{GPT-4} and \texttt{Gemini-1.5-flash-8B} consistently received highest ratings, especially in scenarios in which lecture and/or solution information was included. The Gemini family of models and  \texttt{Pixtral-12b-2409} model showed better robustness in many scenarios and came in second overall. Notably, Gemini models achieved an average accuracy advantage of 5\% over GPT-family models in the "pure" task context (without lecture or solution information). The \texttt{LLaVAModel1.5-7b}, on the other hand, demonstrated the lowest average accuracy scores, particularly in environments without lecture and solution data.

@@

\paragraph{Pre-experiments} We were not successful in processing the dataset with small non-chat models. Those  models would return short meaningless answers, like "Yes" to a "X or Y" question, or "no idea" to a completely defined task with all relevant context included. These were multimodal models below 1B on Huggingface, not tuned to be used in a chat scenario (and prepending a "You are a helpful assistant..." to the beginning of the prompt was not enough to fix this).  We are not sure how to use those models for our purposes as of now.

@@

\paragraph{LLaVa 1.5 7B} The exact model used is: \texttt{llava-hf/llava-1.5-7b-hf} (and there is a larger sister 13b model). It is based on LLaMA and Vicuna, and is fine-tuned on GPT-generated multimodal instruction-following data. It expects inputs as conversation, allowing the preprocessor to convert textual input into format "USER: {request} ASSISTANT: " where the assistant has to fill in the rest.

@@

\begin{table}
\centering
\begin{tabular}{rcc}
	\toprule
	Model name & llava-1.5-7b-hf & \\
	Size & 7B & \\
	Published & 09.2023 & \\
	Type & SLM & \\
	Chat model & yes & \\
	Parent models & LLaMA, Vicuna & \\
	Data & GPT-generated multimodal instruction-following data & \\
	\bottomrule
\end{tabular}
\caption{Model Comparison: Benchmarked Models}
\end{table}

@@

\section{Knowledge Distillation \& Soft Prompting}
\label{knowledge-distillation}

>>>>

\section{Metrics}
We evaluate model's performance with question answering accuracy domain-wise in order to have a fair comparison with leaderboards. The reasoning steps can be evaluated with semantic similarity metrics (e.g. adopted from machine translation) such as BLEU, METEOR, ROUGE, and cosine similarity.
\subsection{Multiple choice Evaluation}
Owing to the simplicity of the test format, only accuracy score is computed, following original evaluation strategy by \cite{lu2022learn}. It was determined that the computation of the overall score should assign the highest weight (\textbf{50\%}) to the ability to provide correct answers (accuracy score), as this represents the most critical aspect of performance evaluation.
\subsection{Answer Reasoning Evaluation}
Due to the pecularity of scientific texts and approaches to the evaluation of automatically generated texts, the following evaluation approaches were chosen. Our validation process is consisted of the metrics introduced in the original paper \cite{lu2022learn}  with a few additional metrics and techniques that we consider of high importance in regards to the scientific textual data.

@@

\label{example:1}
\begin{itemize}
    \item \textbf{Reference:} Plants absorb sunlight to produce energy.
    \item \textbf{Candidate:} Sunlight is absorbed by plants to create energy.
\end{itemize}

\paragraph{METEOR}
METEOR (\cite{banerjee2005meteor}) evaluates text similarity based on precision, recall, and alignment of semantically meaningful components like synonyms and stems, making it well-suited for capturing the nuanced language used in scientific reasoning.
By incorporating a weighted F-score and penalties for excessive mismatches, METEOR provides a balanced evaluation of fluency and accuracy, crucial for assessing LLM-generated answers on the Science QA Dataset. Due to these balanced text evaluation, this metric has \textbf{15\%} impact on the overall score.
% \subsubsection{Cosine similarity with SciBERT}
% We extended the evaluation of cosine similarity using SciBERT (\cite{beltagy2019scibert}) embeddings. These vectors are more consistent with the domain-specific terminology and reasoning patterns present in the Science QA Dataset because it has been properly pre-trained on scientific literature.
% SciBERT generates more accurate similarity embeddings because it better reflects the complex relationships between scientific concepts than general-purpose models like those in Sentence Transformers.
% When assessing the logical and factual coherence of scientific reasoning, the structural and contextual semantics specific to scientific texts are reflected in the embeddings produced by SciBERT.
% Assessments for jobs requiring sophisticated scientific understanding are more reliable when SciBERT is used for cosine similarity since it guarantees that the evaluation metric is domain-specific.

@@

We evaluate: 
\begin{enumerate}
	\item OpenAI models: \texttt{GPT-4}, \texttt{GPT-4o-mini}
	\item MistralAI model: \texttt{Pixtral-12b-2409}
    \item Google models: \texttt{Gemini-1.5-flash}, \texttt{Gemini-1.5-flash-8B}
	\item Llava model:  \texttt{llava-hf/llava-1.5-7b-hf}
\end{enumerate}

\subsection{Results}
% description of metrics, why they were used + analytics of results
\paragraph{Accuracy}
The highest performance was observed in settings that included solutions, indicating that the models were generally capable of extracting relevant information effectively. Among these, settings combining both lecture and solution information typically yielded slightly better results than those with solutions alone. \texttt{GPT-4} and \texttt{Gemini-1.5-flash-8B} consistently received highest ratings, especially in scenarios in which lecture and/or solution information was included. The Gemini family of models and  \texttt{Pixtral-12b-2409} model showed better robustness in many scenarios and came in second overall. Notably, Gemini models achieved an average accuracy advantage of 5\% over GPT-family models in the "pure" task context (without lecture or solution information). The \texttt{LLaVAModel1.5-7b}, on the other hand, demonstrated the lowest average accuracy scores, particularly in environments without lecture and solution data.

@@

\paragraph{Pre-experiments} We were not successful with processing with small non-chat models. For the given text + image, the models would return short non-meaningful answers, like "Yes" to a "X or Y" question, or "no idea" to a completely defined task with all relevant context included. These were multimodal models below 1B on Huggingface, not tuned to be used in a chat scenario (and prepending a "You are a helpful assistant..." to the beginning of the prompt was not enough to fix this). 
We are not sure how to use those models for our purposes as of now.

@@

\paragraph{LLaVa} The exact model used is: \texttt{llava-hf/llava-1.5-7b-hf} (and there is a larger sister 13b model). It is based on LLaMA and Vicuna, and is fine-tuned on GPT-generated multimodal instruction-following data. It expects inputs as conversation, allowing the preprocessor to convert textual input into format "USER: {request} ASSISTANT: " where the assistant has to fill in the rest.

@@

% \begin{table}
% \centering
% \begin{tabular}{rcc}
% 	\toprule
% 	Model name & llava-1.5-7b-hf & \\
% 	Size & 7B & \\
% 	Published & 09.2023 & \\
% 	Type & SLM & \\
% 	Chat model & yes & \\
% 	Parent models & LLaMA, Vicuna & \\
% 	Data & GPT-generated multimodal instruction-following data & \\
% 	\bottomrule
% \end{tabular}
% \caption{Model Comparison: Benchmarked Models}
% \end{table}

% \begin{table}
% \centering
% \begin{tabular}{rcc}
% 	\toprule
% 	Model name & llava-1.5-7b-hf & \\
% 	Size & 7B & \\
% 	Published & 09.2023 & \\
% 	Type & SLM & \\
% 	Chat model & yes & \\
% 	Parent models & LLaMA, Vicuna & \\
% 	Data & GPT-generated multimodal instruction-following data & \\
% 	\bottomrule
% \end{tabular}
% \caption{Model Comparison: Benchmarked Models}
% \end{table}

@@

\section{Knowledge Distillation \& Soft Prompting}
TBD: as mentioned on the slides, we will conduct the described experiments and thoroughly report the results later.

\paragraph{Soft Prompting}
Goal: Investigate how well knowledge distillation works compared to prompt tuning for science reasoning tasks and to zero-shot inference

We freeze a LLM and learn prefix to guide the model
Prompt tuning increases/decreases the performance by {x}%.
This is further compared to Knowledge Distillation from outputs of Teacher model, which performs {insert here} in comparison to prompt tuning

\paragraph{Soft Prompting \& Knowledge Distillation}
Goal: Investigate how well knowledge distillation works compared to prompt tuning for science reasoning tasks.

This is further compared to Knowledge Distillation from outputs of Teacher model, which performs {insert here} in comparison to fine tuning
We try different prompting techniques to obtain useful outputs for learning of the Student
Prompting technique {X} results in the best performance

\paragraph{Soft Prompting \& Knowledge Distillation}
Goal: Investigate if data distilled from Teacher can be (more) useful as compared to using just the training data for science reasoning tasks.
\paragraph{Agentic Setup \& Knowledge Distillation}
Goal: Investigate if data distilled from Teacher + retrieved sources (+ explanation of retrieved data by Teacher) allows us to learn a better Student model. Compare to: Student + RAG capabilities

\label{knowledge-distillation}
